---
title: " Modern Data Mining, HW 4"
author:
- Ludwig Zhao
- Nick Pangakis
- Xinyi Wang
date: '11:59 pm, 03/20, 2021'
output:
  html_document:
    code_folding: show
    highlight: haddock
    number_sections: yes
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '4'
  word_document:
    toc: yes
    toc_depth: '4'
urlcolor: blue
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results =TRUE,include = TRUE, fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output
if(!require('pacman')) {install.packages('pacman')}
pacman::p_load(bestglm, glmnet, leaps, car, tidyverse, pROC, caret) # add the packages needed
```

\pagebreak

# Overview

Logistic regression is used for modeling categorical response variables. The simplest scenario is how to identify risk factors of heart disease? In this case the response takes a possible value of `YES` or `NO`. Logit link function is used to connect the probability of one being a heart disease with other potential risk factors such as `blood pressure`, `cholestrol level`, `weight`. Maximum likelihood function is used to estimate unknown parameters. Inference is made based on the properties of MLE. We use AIC to help nailing down a useful final model. Predictions in categorical response case is also termed as `Classification` problems. One immediately application of logistic regression is to provide a simple yet powerful classification boundaries. Various metrics/criteria are proposed to evaluate the quality of a classification rule such as `False Positive`, `FDR` or `Mis-Classification Errors`. 

LASSO with logistic regression is a powerful tool to get dimension reduction. 


## Objectives

- Understand the model
  - logit function
    + interpretation
  - Likelihood function
- Methods
    - Maximum likelihood estimators
        + Z-intervals/tests
        + Chi-squared likelihood ratio tests
- Metrics/criteria 
    - Sensitivity/False Positive
    - True Positive Prediction/FDR
    - Misclassification Error/Weighted MCE
    - Residual deviance
    - Training/Testing errors

- LASSO 

- R functions/Packages
    - `glm()`, `Anova`
    - `pROC`
    - `cv.glmnet`
  
## R Markdown / Knitr tips

You should think of this R Markdown file as generating a polished report, one that you would be happy to show other people (or your boss). There shouldn't be any extraneous output; all graphs and code run should clearly have a reason to be run. That means that any output in the final file should have explanations.

A few tips:

* Keep each chunk to only output one thing! In R, if you're not doing an assignment (with the `<-` operator), it's probably going to print something.
* If you don't want to print the R code you wrote (but want to run it, and want to show the results), use a chunk declaration like this: `{r, echo=F}`. Notice this is set as a global option. 
* If you don't want to show the results of the R code or the original code, use a chunk declaration like: `{r, include=F}`
* If you don't want to show the results, but show the original code, use a chunk declaration like: `{r, results='hide'}`.
* If you don't want to run the R code at all use `{r, eval = F}`.
* We show a few examples of these options in the below example code. 
* For more details about these R Markdown options, see the [documentation](http://yihui.name/knitr/options/).
* Delete the instructions and this R Markdown section, since they're not part of your overall report.

## Review

Review the code and concepts covered in

* Module Logistic Regressions/Classification
* Module LASSO in Logistic Regression

## This homework

We have two parts in this homework. Part I is guided portion of work, designed to get familiar with elements of logistic regressions/classification. Part II, we bring you projects. You have options to choose one topic among either Credit Risk via LendingClub or Diabetes and Health Management. Find details in the projects. 



# Part I: Framingham heart disease study 

We will continue to use the Framingham Data (`Framingham.dat`) so that you are already familiar with the data and the variables. All the results are obtained through training data. 

Liz is a patient with the following readings: `AGE=50, GENDER=FEMALE, SBP=110, DBP=80, CHOL=180, FRW=105, CIG=0`. We would be interested to predict Liz's outcome in heart disease. 

To keep our answers consistent, use a subset of the data, and exclude anyone with a missing entry. For your convenience, we've loaded it here together with a brief summary about the data.

```{r data preparation, include=F}
# Notice that we hide the code and the results here
# Using `include=F` in the chunk declaration. 
hd_data <- read.csv("Framingham.dat")
str(hd_data) 

### Renames, setting the variables with correct natures...
names(hd_data)[1] <- "HD"
hd_data$HD <- as.factor(hd_data$HD)
hd_data$SEX <- as.factor(hd_data$SEX)
str(hd_data)
#tail(hd_data, 1)    # The last row is for prediction
hd_data.new <- hd_data[1407,] # The female whose HD will be predicted.
hd_data <- hd_data[-1407,]  # take out the last row 
hd_data.f <- na.omit(hd_data)
```

We note that this dataset contains 311 people diagnosed with heart disease and 1095 without heart disease.
```{r table heart disease, echo = F, comment = " "}
# we use echo = F to avoid showing this R code
# notice the usage of comment = " " here in the header
table(hd_data$HD) # HD: 311 of "0" and 1095 "1" 
```

After a quick cleaning up here is a summary about the data:
```{r data summary, comment="     "}
# using the comment="     ", we get rid of the ## in the output.
summary(hd_data.f)

row.names(hd_data.f) <- 1:1393
set.seed(1)
indx <- sample(1393, 5)
hd_data.f[indx, ]
set.seed(1)
hd_data.f[sample(1393, 5), ]
```

## Identify risk factors

### Understand the likelihood function
Conceptual questions to understand the building blocks of logistic regression. All the codes in this part should be hidden. We will use a small subset to run a logistic regression of `HD` vs. `SBP`. 

i. Take a random subsample of size 5 from `hd_data_f` which only includes `HD` and `SBP`. Also set  `set.seed(50)`. List the five observations neatly below. No code should be shown here.

```{r, echo=F}
set.seed(50)

hd_data_sub <- hd_data.f %>% 
  select(HD,SBP) %>% 
  sample_n(5)
head(hd_data_sub)
```


ii. Write down the likelihood function using the five observations above.

\(\mathcal{L}(\beta_0,\beta_1|Data) = Prob(HD=1|SBP=152) \times Prob(HD=0|SBP=110)\times  Prob(HD=0|SBP=154) \times Prob(HD=1|SBP=160)\times  Prob(HD=0|SBP=182) \)

iii. Find the MLE based on this subset using glm(). Report the estimated logit function of `SBP` and the probability of `HD`=1. Briefly explain how the MLE are obtained based on ii. above.

```{r, echo=F}
fit1_sub <- glm(HD~SBP, hd_data_sub, family=binomial(logit))
#summary(fit1_sub, results=TRUE)
```

\(Prob(HD=1|SBP) = \frac{e^{-2.5456 + 0.0140*SBP}}{1 + e^{-2.5456 + 0.0140*SBP}}\)

The MLE are obtained by finding the parameters that maximize the probability of seeing the data as it appears. To do this, we take the log of the likelihood function and then find the minimum of the negative log-likelihood function.

iv. Evaluate the probability of Liz having heart disease. 

```{r, echo=F}
fit1.predict <- predict(fit1_sub, hd_data.new, type="response") 
fit1.predict
```

There is a 24.2 percent chance that Liz has heart disease.

### Identify important risk factors for `Heart.Disease.`

We focus on understanding the elements of basic inference method in this part. Let us start a fit with just one factor, `SBP`, and call it `fit1`. We then add one variable to this at a time from among the rest of the variables. For example
```{r, results='hide'}
fit1 <- glm(HD~SBP, hd_data.f, family=binomial)
summary(fit1)
fit1.1 <- glm(HD~SBP + AGE, hd_data.f, family=binomial)
summary(fit1.1)
# you will need to finish by adding each other variable 
# Sex
fit1.2 <-glm(HD~SBP + SEX, hd_data.f, family=binomial)
summary(fit1.2)
# DBP
fit1.3 <-glm(HD~SBP + DBP, hd_data.f, family=binomial)
summary(fit1.3)
# Chol
fit1.4 <-glm(HD~SBP + CHOL, hd_data.f, family=binomial)
summary(fit1.4)
# FRW
fit1.5 <-glm(HD~SBP + FRW, hd_data.f, family=binomial)
summary(fit1.5)
# CIG
fit1.6 <-glm(HD~SBP + CIG, hd_data.f, family=binomial)
summary(fit1.6)

```

i. Which single variable would be the most important to add?  Add it to your model, and call the new fit `fit2`.  

```{r}
which.min(c(fit1.1[["aic"]],fit1.2[["aic"]],fit1.3[["aic"]],fit1.4[["aic"]],fit1.5[["aic"]],
          fit1.6[["aic"]])) # SEX

fit2 <-glm(HD~SBP + SEX, hd_data.f, family=binomial)
```

Based on the model with the lowest AIC, the single most important variable to add is `SEX`. `SEX` also has the highest $|z|$ value out of all of the variables.

We will pick up the variable either with highest $|z|$ value, or smallest $p$ value. Report the summary of your `fit2` Note: One way to keep your output neat, we will suggest you using `xtable`. And here is the summary report looks like.
```{r the most important addition, results='asis', comment="   "}
## How to control the summary(fit2) output to cut some junk?
## We could use packages: xtable or broom. 
## Assume the fit2 is obtained by SBP + AGE
library(xtable)
options(xtable.comment = FALSE)
print(xtable(fit2),type="html")
```

ii. Is the residual deviance of `fit2` always smaller than that of `fit1`? Why or why not?

Yes, it will always be smaller because we are adding additional variables to the model. 
  
iii. Perform both the Wald test and the Likelihood ratio tests (Chi-Squared) to see if the added variable is significant at the .01 level.  What are the p-values from each test? Are they the same? 

```{r}
# Wald test
summary(fit2) 

# Likelihood ratio test
anova(fit2, test="Chisq")
```

The added variable is significant at the 0.01 level for each model. The p-value for the Wald test is 1.0e-10 and the p-value for the LR test is 3.8e-11. These values are essentially the same since they are very close to zero. 

###  Model building

Start with all variables. Our goal is to fit a well-fitting model, that is still small and easy to interpret (parsimonious).

i. Use backward selection method. Only keep variables whose coefficients are significantly different from 0 at .05 level. Kick out the variable with the largest p-value first, and then re-fit the model to see if there are other variables you want to kick out.

```{r}
fit0 <-glm(HD~. , hd_data.f, family=binomial)
summary(fit0)
fit3.1 <- update(fit0, .~. -DBP)
summary(fit3.1)
fit3.2<- update(fit3.1, .~. -FRW)
summary(fit3.2)
fit3.3<- update(fit3.2, .~. -CIG)
summary(fit3.3)
```


ii. Use AIC as the criterion for model selection. Find a model with small AIC through exhaustive search. Does exhaustive search  guarantee that the p-values for all the remaining variables are less than .05? Is our final model here the same as the model from backwards elimination? 

```{r}
# Get the design matrix without 1's and HD
Xy_design <- model.matrix(HD ~.+0, hd_data.f) 
# Attach y as the last column.
Xy <- data.frame(Xy_design, hd_data.f$HD)   

fit.all <- bestglm(Xy, family = binomial, method = "exhaustive", IC="AIC", nvmax = 10) 
fit.all
```

The exhaustive search does not guarantee that the remaining variables are significant at the 0.05 level. `FRW`, for example, is not significant at the 0.05 level. The final model from the exhaustive search is not the same as the backwards selection. Both `FRW` and `CIG` are included in the exhaustive search, but not in the backwards search. 

iii. Use the model chosen from part ii. as the final model. Write a brief summary to describe important factors relating to Heart Diseases (i.e. the relationships between those variables in the model and heart disease). Give a definition of “important factors”. 

```{r}
fit.all
```

The final model shows that numerous features are associated with increased risk of heart disease. "Important" factors might include features with higher z-values. In this case, higher age, being male, and higher systolic blood pressure are all strongly associated with heart disease. Higher cholesterol, higher weight, and cigarette smoking are associated with heart disease to a lesser degree. 

iv. What is the probability that Liz will have heart disease, according to our final model?

```{r}

fit_final <- glm(HD~AGE+SEX+SBP+CHOL+FRW+CIG, family=binomial, data=hd_data.f)
fit_final.predict <- predict(fit_final, hd_data.new, type="response") 
fit_final.predict
```

There is a 3.46 percent chance that Liz has heart disease.

##  Classification analysis

### ROC/FDR

i. Display the ROC curve using `fit1`. Explain what ROC reports and how to use the graph. Specify the classifier such that the False Positive rate is less than .1 and the True Positive rate is as high as possible.

```{r,warning=F,message=F}
fit1.roc<- roc(hd_data.f$HD, fit1$fitted, plot=T, col="blue")
# Specify the classifier such that the False Positive rate is less than .1 and the True Positive rate is as high as possible.
optimal_sensitivity <- max(fit1.roc[["sensitivities"]][(1 - fit1.roc$specificities) < .1])
(classifier_threshold <- fit1.roc[["thresholds"]][fit1.roc[["sensitivities"]] == optimal_sensitivity])

```

The ROC curve informs how well the model is able to distinguish between classes. The curve shows the proportion of classified true positives (y-axis) and false positives (x-axis). The curve is generated by varying the probability threshold for classification.This curve can be use to select an optimal threshold.

A classification threshold of 0.298 produces a False Positive rate of less than .1 and a True Positive rate as high as possible. 

ii. Overlay two ROC curves: one from `fit1`, the other from `fit2`. Does one curve always contain the other curve? Is the AUC of one curve always larger than the AUC of the other one? Why or why not?

```{r,warning=F,message=F}
fit2.roc<- roc(hd_data.f$HD, fit2$fitted, plot=T, col="blue")

plot(1-fit1.roc$specificities, 
     fit1.roc$sensitivities, col="red", lwd=3, type="l",
     xlab="False Positive", 
     ylab="Sensitivity")
lines(1-fit2.roc$specificities, fit2.roc$sensitivities, col="blue", lwd=3)
legend("bottomright",
       c(paste0("fit1 AUC=", round(fit1.roc$auc,2)), 
         paste0("fit2 AUC=", round(fit2.roc$auc, 2))), 
       col=c("red", "blue"),
       lty=1)
```

The curve in fit2 always contains the curve of fit1 and the AUC curve of fit2 will always be larger than fit1 because fit2 adds additional variables to the model.  

iii.  Estimate the Positive Prediction Values and Negative Prediction Values for `fit1` and `fit2` using .5 as a threshold. Which model is more desirable if we prioritize the Positive Prediction values?

```{r}
fit1_predictions <- as.numeric(ifelse(fit1$fitted > 1/2, "1", "0"))
(confusion_matrix_fit1 <- table(fit1_predictions,hd_data.f$HD))
positive_pred_fit1 <- confusion_matrix_fit1[2,2] / sum(confusion_matrix_fit1[2,])
paste("Positive Prediction Values Fit1:",round(positive_pred_fit1,2))
negative_pred_fit1 <- confusion_matrix_fit1[1,1] / sum(confusion_matrix_fit1[1,])
paste("Negative Prediction Values Fit1:",round(negative_pred_fit1,2))

fit2_predictions <- as.numeric(ifelse(fit2$fitted > 1/2, "1", "0"))
(confusion_matrix_fit2 <- table(fit2_predictions,hd_data.f$HD))
positive_pred_fit2 <- confusion_matrix_fit2[2,2] / sum(confusion_matrix_fit2[2,])
paste("Positive Prediction Values Fit2:",round(positive_pred_fit2,2))
negative_pred_fit2 <- confusion_matrix_fit2[1,1] / sum(confusion_matrix_fit2[1,])
paste("Negative Prediction Values Fit2:",round(negative_pred_fit2,2))
```

Looking at the above confusion matrices, if we want to prioritize the positive predictive values, then we should choose fit2 because the positive prediction rate is higher for fit2 (.47) than fit1 (.45).

iv.  For `fit1`: overlay two curves,  but put the threshold over the probability function as the x-axis and positive prediction values and the negative prediction values as the y-axis.  Overlay the same plot for `fit2`. Which model would you choose if the set of positive and negative prediction values are the concerns? If you can find an R package to do so, you may use it directly.


```{r,warning=F,message=F}
fit1.roc<- roc(hd_data.f$HD, fit1$fitted)
fit2.roc<- roc(hd_data.f$HD, fit2$fitted)

coordinates_fit1 <- coords(fit1.roc, x = "all", input = "threshold", ret = c("threshold", "ppv","npv"))
coordinates_fit2 <- coords(fit2.roc, x = "all", input = "threshold", ret = c("threshold", "ppv","npv"))

coordinates_fit1$fit1 <- 1
coordinates_fit2$fit1 <- 0

coordinates <- rbind(coordinates_fit1,coordinates_fit2)

coordinates %>% 
  ggplot() +
  geom_point(aes(x=threshold,y=ppv,color=as.factor(fit1))) +
  geom_point(aes(x=threshold,y=npv,color=as.factor(fit1))) +
  scale_x_continuous(name = "Classification Thresholds") +
  scale_y_continuous(name = "Prediction Values") +
  scale_color_discrete(name = "Model", labels = c("Fit2", 
                                                  "Fit1")) +
  geom_label(
    label="Negative Prediction Values", 
    x=.5,
    y=.85,
    label.padding = unit(0.1, "lines"), # Rectangle size around label
    label.size = NA,
    color = "black",
    fill="#56B4E9",
    size = 4
  ) +
  geom_label(
    label="Positive Prediction Values", 
    x=.5,
    y=.33,
    label.padding = unit(0.1, "lines"), # Rectangle size around label
    label.size = NA,
    color = "black",
    fill="#56B4E9",
    size = 4
  ) 

```

Looking at the above plot, if the set of positive and negative prediction values was the main criteria for selecting the model, then I would select fit2 because the positive and negative prediction values appear to be slightly higher overall for fit2.

### Cost function/ Bayes Rule

Bayes rules with risk ratio $\frac{a_{10}}{a_{01}}=10$ or $\frac{a_{10}}{a_{01}}=1$. Use your final model obtained from Part 1 to build a class of linear classifiers.

i.  Write down the linear boundary for the Bayes classifier if the risk ratio of $a_{10}/a_{01}=10$.

\(\hat P(Y=1 \vert x) > \frac{0.1}{(1+0.1)}=0.0909 \)

ii. What is your estimated weighted misclassification error for this given risk ratio?

```{r}
fit_final.pred.bayes <- as.factor(ifelse(fit_final$fitted > 0.0909, "1", "0"))
MCE.bayes <- (10*sum(fit_final.pred.bayes[hd_data.f$HD == "1"] != "1")
              + sum(fit_final.pred.bayes[hd_data.f$HD == "0"] != "0"))/length(hd_data.f$HD)
MCE.bayes
```

iii.  How would you classify Liz under this classifier?

```{r}
fit_final.predict <- predict(fit_final, hd_data.new, type="response") 
(liz_classification <- as.factor(ifelse(fit_final.predict > 0.0909, "1", "0")))
```

Under the the new threshold, Liz would still be classified as a 0 (i.e., no heart disease).

iv. Bayes rule gives us the best rule if we can estimate the probability of `HD-1` accurately. In practice we use logistic regression as our working model. How well does the Bayes rule work in practice? We hope to show in this example it works pretty well.

Now, draw two estimated curves where x = threshold, and y = misclassification errors, corresponding to the thresholding rule given in x-axis.

v. Use weighted misclassification error, and set $a_{10}/a_{01}=10$. How well does the Bayes rule classifier perform? 

```{r}
thresholds <- seq(0,1,.01)
mat <- matrix(NA, nrow = length(thresholds), ncol = 3)
for (i in 1:length(thresholds)){
  fit_final.pred.bayes <- as.factor(ifelse(fit_final$fitted > thresholds[i], "1", "0"))
  MCE.bayes_alpha10 <- (10*sum(fit_final.pred.bayes[hd_data.f$HD == "1"] != "1") +
                          sum(fit_final.pred.bayes[hd_data.f$HD == "0"] != "0"))/length(hd_data.f$HD)
  MCE.bayes_alpha1 <- (sum(fit_final.pred.bayes[hd_data.f$HD == "1"] != "1") +
                          sum(fit_final.pred.bayes[hd_data.f$HD == "0"] != "0"))/length(hd_data.f$HD)
  mat[i,1] <- thresholds[i]
  mat[i,2] <- MCE.bayes_alpha1
  mat[i,3] <- MCE.bayes_alpha10
}

colnames(mat) <- c("Thresholds","MCE_alpha1","MCE_alpha10")
mat <- data.frame(mat)


mat %>% 
  ggplot() +
  geom_point(aes(x=Thresholds,y=MCE_alpha10)) +
  scale_x_continuous(name = "Classification Thresholds") +
  scale_y_continuous(name = "Misclassification Error")


```


vi. Use weighted misclassification error, and set $a_{10}/a_{01}=1$. How well does the Bayes rule classifier perform? 


```{r}
mat %>% 
  ggplot() +
  geom_point(aes(x=Thresholds,y=MCE_alpha1)) +
  scale_x_continuous(name = "Classification Thresholds") +
  scale_y_continuous(name = "Misclassification Error")
```

Looking at the two curves, the WMCE when $a_{10}/a_{01}=1$ performs notably better, which makes sense because the cost is not significantly higher.

# Part II: Project

## Project Option 1 Credit Risk via LendingClub

## Project Opetion 2  Diabetes and Health Management

