---
title: " Modern Data Mining, HW 1"
author:
- Ludwig Zhao
- Nick Pangakis
- Xinyi Wang
date: 'Due: 11:59PM,  Jan. 30th, 2021'
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '4'
    latex_engine: xelatex
  html_document:
    code_folding: show
    highlight: haddock
    number_sections: yes
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '4'
urlcolor: blue
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "hide", fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output
# check if you have ISLR package, if not, install it
if(!require('pacman')) {install.packages('pacman')}
pacman::p_load(ISLR, readxl, tidyverse, magrittr, dplyr, ggplot2)
if(!require('reshape2')) {install.packages('reshape2')}


```

\pagebreak

# Overview

This is a fast-paced course that covers a lot of material. There will be a large amount of references. You may need to do your own research to fill in the gaps in between lectures and homework/projects. It is impossible to learn data science without getting your hands dirty. Please budget your time evenly. Last-minute work ethic will not work for this course. 

Homework in this course is different from your usual homework assignment as a typical student. Most of the time, they are built over real case studies.  While you will be applying methods covered in lectures, you will also find that extra teaching materials appear here.  The focus will be always on the goals of the study, the usefulness of the data gathered, and the limitations in any conclusions you may draw. Always try to challenge your data analysis in a critical way. Frequently, there are no unique solutions. 

Case studies in each homework can be listed as your data science projects (e.g. on your CV) where you see fit. 



## Objectives 

- Get familiar with `R-studio` and `RMarkdown`
- Hands-on R 
- Learn data science essentials 
    - gather data
    - clean data
    - summarize data 
    - display data
    - conclusion
- Packages
    - `dplyr`
    - `ggplot`

##  Instructions

- **Homework assignments can be done in a group consisting of up to three members**. Please find your group members as soon as possible and register your group on our Canvas site.

- **All work submitted should be completed in the R Markdown format.** You can find a cheat sheet for R Markdown [here](https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf). For those who have never used it before, we urge you to start this homework as soon as possible. 

- **Submit the following files, one submission for each group:**  (1) Rmd file, (2) a compiled PDF or HTML version, and (3) all necessary data files if different from our source data. You may directly edit this .rmd file to add your answers. If you intend to work on the problems separately within your group, compile your answers into one Rmd file before submitting. We encourage that you at least attempt each problem by yourself before working with your teammates. Additionally, ensure that you can 'knit' or compile your Rmd file. It is also likely that you need to configure Rstudio to properly convert files to PDF. [**These instructions**](http://kbroman.org/knitr_knutshell/pages/latex.html#converting-knitrlatex-to-pdf) might be helpful.

- In general, be as concise as possible while giving a fully complete answer to each question. All necessary datasets are available in this homework folder on Canvas. Make sure to document your code with comments (written on separate lines in a code chunk using a hashtag `#` before the comment) so the teaching fellows can follow along. R Markdown is particularly useful because it follows a 'stream of consciousness' approach: as you write code in a code chunk, make sure to explain what you are doing outside of the chunk. 

- A few good or solicited submissions will be used as sample solutions. When those are released, make sure to compare your answers and understand the solutions.


## Review materials

- Study Advanced R Tutorial (to include `dplyr` and `ggplot`)
- Study lecture 1: Data Acquisition and EDA


# Case study 1: Audience Size

How successful is the Wharton Talk Show [Business Radio Powered by the Wharton School](https://businessradio.wharton.upenn.edu/)  


**Background:** Have you ever listened to [SiriusXM](https://www.siriusxm.com/)? Do you know there is a **Talk Show** run by Wharton professors in Sirius Radio?  Wharton launched a talk show called [Business Radio Powered by the Wharton School](https://businessradio.wharton.upenn.edu/) through the Sirius Radio station in January of 2014. Within a short period of time the general reaction seemed to be overwhelmingly positive. To find out the audience size for the show, we designed a survey and collected a data set via MTURK in May of 2014. Our goal was to **estimate the audience size**. There were 51.6 million Sirius Radio listeners then. One approach is to estimate the proportion of the Wharton listeners to that of the Sirius listeners, $p$, so that we will come up with an audience size estimate of approximately 51.6 million times $p$. 

To do so, we launched a survey via Amazon Mechanical Turk ([MTurk](https://www.mturk.com/)) on May 24, 2014 at an offered price of \$0.10 for each answered survey.  We set it to be run for 6 days with a target maximum sample size of 2000 as our goal. Most of the observations came in within the first two days. The main questions of interest are "Have you ever listened to Sirius Radio" and "Have you ever listened to Sirius Business Radio by Wharton?". A few demographic features used as control variables were also collected; these include Gender, Age and Household Income.  

We requested that only people in United States answer the questions. Each person can only fill in the questionnaire once to avoid duplicates. Aside from these restrictions, we opened the survey to everyone in MTurk with a hope that the sample would be more randomly chosen. 

The raw data is stored as `Survey_results_final.csv` on Canvas.

## Data preparation

i. We need to clean and select only the variables of interest. 

Select only the variables Age, Gender, Education Level, Household Income in 2013, Sirius Listener?, Wharton Listener? and Time used to finish the survey.

Change the variable names to be "age", "gender", "education", "income", "sirius", "wharton", "worktime".

```{r data wrangling, echo = TRUE, include=TRUE, warning = FALSE}
radio <- read.csv("data/Survey_results_final.csv")
#change names
radio <- radio %>% 
  select(Answer.Age, Answer.Gender,Answer.Education,Answer.HouseHoldIncome,
         Answer.Sirius.Radio, Answer.Wharton.Radio, WorkTimeInSeconds)
radio <- radio %>% 
  rename(age = 'Answer.Age',
         gender = 'Answer.Gender',
         education = 'Answer.Education',
         income = 'Answer.HouseHoldIncome',
         sirius = 'Answer.Sirius.Radio',
         wharton = 'Answer.Wharton.Radio',
         worktime = 'WorkTimeInSeconds')
```

ii. Handle missing/wrongly filled values of the selected variables

As in real world data with user input, the data is incomplete, with missing values, and has incorrect responses. There is no general rule for dealing with these problems beyond “use common sense.” In whatever case, explain what the problems were and how you addressed them. Be sure to explain your rationale for your chosen methods of handling issues with the data. Do not use Excel for this, however tempting it might be.

Tip: Reflect on the reasons for which data could be wrong or missing. How would you address each case? For this homework, if you are trying to predict missing values with regression, you are definitely overthinking. Keep it simple.

First by looking at number NAs, it seems like there are no NA entries. Diving into each variable, there seems to be some wrong and empty entries in the form of "". 
- Age has several character entries that we transformed to numeric numbers of age. Other demographic variables had different number of empty entries, which were deleted because the numbers were small enough to ignore. 
- Sirius and wharton each had 5 and 4 empty entries, which were deleted. 
- By looking at worktime, those who finished the survey below 10 seconds and above 60 seconds seem to be outliers of worktime. Therefore these entries were deleted as well.
```{r, echo = TRUE, include=TRUE, warning = FALSE,results=TRUE}
sapply(radio, function(x) sum(is.na(x)))
#no actual NA values in each column
#now check what entries are there in age
radio %>% 
  group_by(age) %>% 
  summarize(n = n())
#One entry of 27`, one entry Eighteen (18), one entry female, one entry 223, one empty entry, one entry is 4
radio$age[radio$age == "27`"] <- 27
radio$age[radio$age == "Eighteen (18)"] <- 18
radio <- radio[!radio$age == "223", ]
radio <- radio[!radio$age == "4", ]
radio <- radio[!radio$age == "female", ]
radio <- radio[!radio$age == "", ] 

#now check what entries are there in gender
radio %>% 
  group_by(gender) %>% 
  summarize(n = n())
#There are 6 empty entries, delete the empty entries
#now check what entries are there in education
radio %>% 
  group_by(education) %>% 
  summarize(n = n())
#19 select one

#now check what entries are there in income
radio %>% 
  group_by(income) %>% 
  summarize(n = n())
# 6 empty entries

#now check what entries are there in sirius
radio %>% 
  group_by(sirius) %>% 
  summarize(n = n())
# 5 empty entries

#now check what entries are there in wharton
radio %>% 
  group_by(wharton) %>% 
  summarize(n = n())
# 4 empty entries

#now check what entries are there in worktime
radio %>% 
  group_by(worktime) %>% 
  summarize(n = n())
# no empty entries, but some very short ones and very long ones look like outliers
#delete the empty entries
radio <- radio[!radio$gender == "",]
radio <- radio[!radio$education == "select one",]
radio <- radio[!radio$income == "",]
radio <- radio[!radio$sirius == "",]
radio <- radio[!radio$wharton == "",]
#delete outliers of worktime
ggplot(radio) + 
  geom_histogram(aes(x = worktime), fill = "blue")
radio <- radio %>% 
  filter(worktime > 10 & worktime < 60)
ggplot(radio) + 
  geom_histogram(aes(x = worktime), fill = "blue")
```


iii. Brief summary 

Write a brief report to summarize all the variables collected. Include both summary statistics (including sample size) and graphical displays such as histograms or bar charts where appropriate. Comment on what you have found from this sample. (For example - it's very interesting to think about why would one work for a job that pays only 10cents/each survey? Who are those survey workers? The answer may be interesting even if it may not directly relate to our goal.)

From the summary statistics and graphs, we can see that this sample of participants are mostly in the age between their 20s to 40s. There are more male than female participants where male make up of 57.5% of the sample and female make up of 42.5% of the sample. Most participants either held a bachelor's degree or Some college, no diploma; or Associate’s degree. Few participants held higher than Bachelor's degree and very few held high school graduate or lower education level. Participant's income are relatively evenly spread between 15k-30k, 30k-50k, 30k-75k, and 75k-150k. Few participants' income level is below 15k, and very few above 150k. Overall, this is a group of relatively young and middle aged participants who hold relatively lower to middle socioeconomic status.

```{r, echo = TRUE, include=TRUE, warning = FALSE,results=TRUE}
#summary statistics
summary(radio)
skimr::skim(radio)
#histogram for age
radio$age <- as.numeric(radio$age)
ggplot(radio) + 
  geom_histogram(aes(x = age), fill = "dark blue") +
  labs( title = "Histogram of Age", x = "Age" , y = "Frequency")
#gender
ggplot(radio) +
  geom_bar(aes(x = gender), fill = "dark blue")
radio %>%
  group_by(gender) %>% 
  summarise(cnt = n()) %>% 
  mutate(freq = cnt / sum(cnt)) %>% 
  arrange(desc(freq))
#education
ggplot(radio) +
  geom_bar(aes(x = education), fill = "dark blue")
radio %>%
  group_by(education) %>% 
  summarise(cnt = n()) %>% 
  mutate(freq = cnt / sum(cnt)) %>% 
  arrange(desc(freq))
#bar chart for income
ggplot(radio) +
  geom_bar(aes(x = income), fill = "dark blue")
radio %>%
  group_by(income) %>% 
  summarise(cnt = n()) %>% 
  mutate(freq = cnt / sum(cnt)) %>% 
  arrange(desc(freq))
#bar chart for sirius 
ggplot(radio) +
  geom_bar(aes(x = sirius), fill = "dark blue")
radio %>%
  group_by(sirius) %>% 
  summarise(cnt = n()) %>% 
  mutate(freq = cnt / sum(cnt)) %>% 
  arrange(desc(freq))
#bar chart for wharton
ggplot(radio) +
  geom_bar(aes(x = wharton), fill = "dark blue")
radio %>%
  group_by(wharton) %>% 
  summarise(cnt = n()) %>% 
  mutate(freq = cnt / sum(cnt)) %>% 
  arrange(desc(freq))
```


## Sample properties

The population from which the sample is drawn determines where the results of our analysis can be applied or generalized. We include some basic demographic information for the purpose of identifying sample bias, if any exists. Combine our data and the general population distribution in age, gender and income to try to characterize our sample on hand.

i. Does this sample appear to be a random sample from the general population of the USA?
  This sample appears to have slightly less female, much more younger population in their 20s, and more population who's income is lower than 50k a year compare to the general population of the USA. Overall, this does not appear to be a random sample from the general population
ii. Does this sample appear to be a random sample from the MTURK population?
  This sample appears to have less female, much more younger population in their 20s, almost equal income distribution compare to the MTurk population. Overall, this does not appear to be a random sample from the MTurk population 
Note: You can not provide evidence by simply looking at our data here. For example, you need to find distribution of education in our age group in US to see if the two groups match in distribution. You may need to gather some background information about the MTURK population to have a slight sense if this particular sample seem to a random sample from there... Please do not spend too much time gathering evidence. 

```{r, echo = TRUE, include=TRUE, warning = FALSE,results=TRUE}
#try to look at gender, age, education, and income distribution of our dataset based on both of american population + mturk
#https://www.cloudresearch.com/resources/blog/who-uses-amazon-mturk-2020-demographics/
#Gender
gender_per <- radio %>%
  group_by(gender) %>% 
  summarise(cnt = n()) %>% 
  mutate(radio = cnt / sum(cnt))
gender_per <- gender_per[-2]
gender_per <- gender_per %>% 
  mutate(MTURK = c(0.57, 0.43),
         national = c(0.51, 0.49))
gender_per <- melt(gender_per)
ggplot(gender_per, aes(x = gender, y = value, fill = factor(variable))) +
  geom_bar(stat = "identity",position = position_dodge())+
  ggtitle("Gender Distribution of Radio Survey, MTURK, and National")
#Age
radio <- radio %>% 
  mutate(agegroup = case_when(age >= 18 & age <= 29 ~ '20s',
                              age >= 30 & age <= 39 ~ '30s',
                              age >= 40 & age <= 49 ~ '40s',
                              age >= 50 & age <= 59 ~ '50s',
                              age >= 60 ~ 'over 60s'))

age_per <- radio %>%
  group_by(agegroup) %>%
  summarise(cnt = n()) %>% 
  mutate(radio = cnt / sum(cnt))
age_per <- age_per[-2]
age_per <- age_per %>% 
  mutate(MTURK = c(0.297, 0.368, 0.168, 0.107, 0.06),
         national = c(0.15, 0.165, 0.156, 0.172, 0.356))
age_per <- melt(age_per)

ggplot(age_per, aes(x = agegroup, y = value, fill = factor(variable))) +
  geom_bar(stat = "identity",position = position_dodge())+
  ggtitle("Age Distribution of Radio Survey, MTURK, and National")
#Income
income_per <- radio %>%
  group_by(income) %>% 
  summarise(cnt = n()) %>% 
  mutate(radio = cnt / sum(cnt))
income_per <- income_per[-2]
income_per <- income_per %>% 
  mutate(MTURK = c(0.182, 0.218, 0.264, 0.221, 0.05, 0.06),
         national = c(0.172, 0.165, 0.202, 0.242, 0.15, 0.6))
income_per <- melt(income_per)
ggplot(income_per, aes(x = income, y = value, fill = factor(variable))) +
  geom_bar(stat = "identity",position = position_dodge())+
  ggtitle("Income Distribution of Radio Survey, MTURK, and National")

```


## Final estimate

Give a final estimate of the Wharton audience size in January 2014. Assume that the sample is a random sample of the MTURK population, and that the proportion of Wharton listeners vs. Sirius listeners in the general population is the same as that in the MTURK population. Write a brief executive summary to summarize your findings and how you came to that conclusion.

To be specific, you should include:

1. Goal of the study
2. Method used: data gathering, estimation methods
3. Findings
4. Limitations of the study. 

**Executive Summary of Wharton Radio MTurk Survey**

### Goal of the study
  The goal of this study was to find out the audience size for the Wharton radio show in 2014. To estimate the audience size, a survey via Amazon Mechanical Turk was administered. The goal of the maximum sample was 2000 people during a 6 days span.  The main questions of interest are "Have you ever listened to Sirius Radio" and "Have you ever listened to Sirius Business Radio by Wharton?". A few demographic features used as control variables were also collected; these include Gender, Age and Household Income.  

### Estimation Methods

In the existing survey data, we have 4 independent demographic varibles `income`, `gender`, `education` and `age`. To study the significance of each variable on the probability of listening Wharton Talk Show ($P(X)$), we performed a logistic regression as follow 

```{r,echo = TRUE, include=TRUE, warning = FALSE,results=TRUE}
radio$wharton <- as.factor(radio$wharton)
radio$education <- as.factor(radio$education)
radio$income <- as.factor(radio$income)
radio$gender <- as.factor(radio$gender)
summary(glm(wharton ~ income + education + gender + age , data = radio, family = "binomial"))
```

The summary indicates that the `gender` and `education` with Graduate or professional degree have significant effects on the outcome ($P(X)$).

Therefore, we proposed following model

$$
P(X) = P(X|M \cap G)P(M \cap G) + P(X|\bar{M} \cap G)P(\bar{M} \cap G) + P(X|M \cap \bar{G})P(M \cap \bar{G}) + P(X|\bar{M} \cap \bar{G})P(\bar{M} \cap \bar{G})
$$
Where $X$ is listnting Wharton Talk Show on Sirius MX, $M$ is male, $G$ is having Graduate or professional degree.

However, due to lack of information on percentage cross demographics on a national sample (i.e. percentage of male who have graduate or professional degree), we have to simplify the model into the following that only includes gender 
$$
P(X) = P(X|M)P(M) + P(X|\bar{M})P(\bar{M})
$$

```{r, echo = TRUE, include=TRUE, warning = FALSE,results=TRUE}
P_XM <- radio %>% 
  filter(gender == "Male" & sirius == "Yes") %>%
  group_by(wharton) %>% 
  summarise(cnt = n()) %>% 
  mutate(freq = cnt / sum(cnt)) %>% 
  filter(wharton == "Yes") %>% 
  select(freq)
  
P_XNM <- radio %>% 
  filter(gender == "Female" & sirius == "Yes") %>%
  group_by(wharton) %>% 
  summarise(cnt = n()) %>% 
  mutate(freq = cnt / sum(cnt)) %>% 
  filter(wharton == "Yes") %>% 
  select(freq)
  
P_M <- 0.43 # percentage of male participants on MTurk 
P_NM <- 0.57 # percentage of female participants on MTurk
P_total <- P_XM*P_M + P_XNM * P_NM
Estimate = P_total$freq * 51.6
Estimate # in millions
```

### Findings
  Based on our sample and the estimation method, we found that Wharton listeners make up of 4.24% of

### Limitations
  There are several limitations to this study. First, the study participants were not representative of a nationally . Second, only one question was used to estimate whether the participant is a listener of the Wharton radio show. Other related questions such as listening frequency, whether the participants listen to other similar shows could be asked to better understand the audience. 

## New task

Now suppose you are asked to design a study to estimate the audience size of Wharton Business Radio Show as of today: You are given a budget of $1000. You need to present your findings in two months. 

Write a proposal for this study which includes:

1. Method proposed to estimate the audience size.
2. What data should be collected and where it should be sourced from.
Please fill in the google form to list your platform where surveys will be launched and collected [HERE](https://forms.gle/8SmjFQ1tpqr6c4sa8) 

A good proposal will give an accurate estimation with the least amount of money used. 

Our proposed method of estimating audience size is to use the simple logic of ratios to approximate the number of Wharton listeners. Prop of Sirius listeners in sample/Prop of Wharton listeners in sample = Number of Sirius listeners in population/Number of Wharton listeners in population. 

If we continue the assumption that the proportion of Wharton listeners vs. Sirius listeners in the general population is the same as that in the MTURK population, then we should run another survey on MTURK to estimate Wharton audience size.

State the name and Attach a URL where you want to launch a survey or describe your own way to collect information
https://blog.mturk.com/editing-the-survey-link-project-template-in-the-ui-7c75285105fb

Pros: cheap and quick way to survey a high number of respondents
Cons: Potentially lower quality data (e.g., professional survey takers, unrepresentative, etc)

# Case study 2: Women in Science


Are women underrepresented in science in general? How does gender relate to the type of educational degree pursued? Does the number of higher degrees increase over the years? In an attempt to answer these questions, we assembled a data set (`WomenData_06_16.xlsx`) from [NSF](https://ncses.nsf.gov/pubs/nsf19304/digest/field-of-degree-women) about various degrees granted in the U.S. from 2006 to 2016. It contains the following variables: Field (Non-science-engineering (`Non-S&E`) and sciences (`Computer sciences`, `Mathematics and statistics`, etc.)), Degree (`BS`, `MS`, `PhD`), Sex (`M`, `F`), Number of degrees granted, and Year.

Our goal is to answer the above questions only through EDA (Exploratory Data Analyses) without formal testing. We have provided sample R-codes in the appendix to help you if needed. 


## Data preparation  

1. Understand and clean the data

Notice the data came in as an Excel file. We need to use the package `readxl` and the function `read_excel()` to read the data `WomenData_06_16.xlsx` into R. 


i. Read the data into R.

```{r read data, echo = TRUE, include=TRUE, warning = FALSE,results=TRUE}
wsci <- read_excel("data/WomenData_06_16.xlsx")
```

ii. Clean the names of each variables. (Change variable names to  `Field`,`Degree`, `Sex`, `Year` and `Number` )

```{r change names,echo = TRUE, include=TRUE, warning = FALSE,results=TRUE}
wsci <- wsci %>%
  rename(Field = 'Field and sex',
         Number = 'Degrees Awarded')
```

iii. Set the variable natures properly. 

```{r factorization,echo = TRUE, include=TRUE, warning = FALSE,results=TRUE}
wsci <- wsci %>%
  mutate(Field   = as.factor(Field),
         Degree  = as.factor(Degree),
         Sex     = as.factor(Sex))
```

iv. Any missing values?
```{r check missing vals,echo = TRUE, include=TRUE, warning = FALSE,results=TRUE}
sapply(wsci, function(x) length(x[is.na(wsci)]))
```
There is no missing values.

2. Write a summary describing the data set provided here. 

```{r summary,echo = TRUE, include=TRUE, warning = FALSE,results=TRUE}
skimr::skim(wsci)
hist(wsci$Year)
```
i. How many fields are there in this data?
There are 5 fields.

ii. What are the degree types? 
There are are 3 degree types

iii. How many year's statistics are being reported here? 
The data are reported from 2006 to 2016, 11 years in total


## BS degrees in 2015

Is there evidence that more males are in science-related fields vs `Non-S&E`? Provide summary statistics and a plot which shows the number of people by gender and by field. Write a brief summary to describe your findings.

```{r, echo = TRUE, include=TRUE, warning = FALSE,results=TRUE}
res <- wsci %>%
  mutate(SE = ifelse(Field != "Non-S&E", "S&E", "Non-S&E")) %>%
  filter(Year == 2015 & Degree == "BS") %>%
  group_by(SE, Sex) %>%
  summarise(n = sum(Number))

res

res %>%
ggplot(aes(x = Sex, y = n, fill = SE)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.y = element_text(angle = 60)) +
  ggtitle("Degrees granted by S&E vs non-S&E by gender") + 
  ylab("Number")+ xlab("")
```

In 2015, there are both more male and female within `Non S&E` than in science-related fields. However, the difference is signficantly lower for male compared to female.



Describe the number of people by type of degree, field, and gender. Do you see any evidence of gender effects over different types of degrees? Again, provide graphs to summarize your findings.

```{r, echo = TRUE, include=TRUE, warning = FALSE,results=TRUE}
res <- wsci %>%
  mutate(SE = ifelse(Field!="Non-S&E" , "S&E", "Non-S&E")) %>%
  filter(Year == 2015) %>%
  group_by(SE, Sex, Degree) %>%
  summarise(SE_number = sum(Number))

res

res %>%
  ggplot(aes(x = Sex, y = SE_number)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(SE~Degree, scales = "free_y") +
  ggtitle("Degrees granted by sex, degree and SE") +
  xlab("") + ylab("Number")
```

The table and the graph above show the number of people by type of degree, field, and gender. 
To summarize, within science & engineering related fields, there are approximately similar numbers of B.S. degree granted to female and male in 2015. However, there are more advanced degrees (M.S. and Ph.D.) granted to male within the field. 
By contrast, within non science & engineering related fields, more degrees are awarded to female regardless of the types of the degree.

We can see gender effect on degree for both science-related fields and non science & engineering related fields, especially the latter one, as stated in the previous paragraph.

## EDA bring all variables 

In this last portion of the EDA, we ask you to provide evidence numerically and graphically: Do the number of  degrees change by gender, field, and time? 


```{r,echo = TRUE, include=TRUE, warning = FALSE,results=TRUE}

res <- wsci %>%
  mutate(SE = ifelse(Field!="Non-S&E" , "S&E", "Non-S&E")) %>%
  group_by(SE, Sex, Year) %>%
  summarise(SE_number = sum(Number)) 
res
res %>%
  ggplot(aes(x = Year, y = SE_number, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(~SE, scales = "free_y") +
  ggtitle("Degrees granted by sex, degree and SE")
```

Without doing any statistical analysis, we can still identify following trends over time:
  + More degrees are granted over the time in both fields and for both and female.
  + The portion of female degree awardees over the total degree awardees grows over the year within non S&E fields.
  + The portion of male degree awardees over the total degree awardees grows over the year within S&E fields.

## Women in Data Science

Finally, is there evidence showing that women are underrepresented in data science? Data science is an interdisciplinary field of computer science, math, and statistics. You may include year and/or degree.

```{r,echo = TRUE, include=TRUE, warning = FALSE,results=TRUE}
wsci %>%
  filter(Field == "Mathematics and statistics" | Field == "Computer sciences") %>%
  group_by(Sex, Year, Degree) %>%
  summarize(n = sum(Number)) %>%
  group_by(Year, Degree) %>%
  mutate(ratio = n / sum(n)) %>%
  filter(Sex == "Female") %>%
  ggplot(aes(x = Year, y = ratio))  + 
  geom_point(aes(colour = Degree)) + geom_line(aes(colour = Degree)) +
  xlab("Female Ratio")
```

The figure above shows that all types of degree over past 10 years, female are significantly underrepresented in the data science field; while around 1/3 of the master degrees are awarded to female, roughly only 1/4 of the Ph.D. and B.S. degrees are awarded to female.

## Final brief report

Summarize your findings focusing on answering the questions regarding if we see consistent patterns that more males pursue science-related fields. Any concerns with the data set? How could we improve on the study?

Based on the analysis, although  more degrees are granted to both male and female either fields, the portion of female science-related field degree awardees are always lower than male ones. This highlighted the underrepresention of female in science and engineering related fields over the past 10 years. Furthermore, this gap is increasing over the year as suggested in ``EDA bring all variables`` section.

In this data set, for each year, there are always more female degree awardees compare to the male ones. In other words, the data set has more female samples compare to male samples. While this could be true, one improvement is that: instead of arbitrarily sampling, the ratio between the number of degrees awarded to female and male for a given year should reflect the actual demographics in the graduating population.

# Case study 3: Major League Baseball

We would like to explore how payroll affects performance among Major League Baseball teams. The data is prepared in two formats record payroll, winning numbers/percentage by team from 1998 to 2014. 

Here are the datasets:

-`MLPayData_Total.csv`: wide format
-`baseball.csv`: long format

```{r, include=TRUE,echo=TRUE}
# Load in data
datapay <- read_csv("data/MLPayData_Total.csv")
baseball <- read_csv("data/baseball.csv")
```

Feel free to use either dataset to address the problems. 

## EDA: Relationship between payroll changes and performance

Payroll may relate to performance among ML Baseball teams. One possible argument is that what affects this year's performance is not this year's payroll, but the amount that payroll increased from last year. Let us look into this through EDA. 

Create increment in payroll

i. To describe the increment of payroll in each year there are several possible approaches. Take 2013 as an example:

    - option 1: diff: payroll_2013 - payroll_2012
    - option 2: log diff: log(payroll_2013) - log(payroll_2012)

Explain why the log difference is more appropriate in this setup.

```{r, include=TRUE,echo=TRUE,results=TRUE}
# Summary data
summary(datapay$p2013)

# Histogram of lagged values
baseball %>%
  group_by(team) %>%
  mutate(diff_lag = payroll - lag(payroll)) %>%
  drop_na() %>%
  ggplot() +
  geom_histogram(aes(x = diff_lag), bins = 30, fill = "blue") +
  labs(title = "Histogram of Yearly Change in Payroll", x = "Change in Payroll" , y = "Frequency")
# Histogram of lagged values (log)
baseball %>%
  group_by(team) %>%
  mutate(diff_lag = log(payroll) - log(lag(payroll))) %>%
  drop_na() %>%
  ggplot() +
  geom_histogram(aes(x = diff_lag), bins = 30, fill = "blue") +
  labs(title = "Histogram of Yearly Change in Payroll (log)", x = "Change in Payroll (log)" , y = "Frequency")
```

A logarithmic transformation is more appropriate in this set up to handle skewness in the data. Looking at the two figures above, the second histogram (on the log scale) appears to approximate a normal distribution closer than the first histogram. 

ii. Create a new variable `diff_log=log(payroll_2013) - log(payroll_2012)`. Hint: use `dplyr::lag()` function.

iii. Create a long data table including: team, year, diff_log, win_pct

```{r, include=TRUE,echo=TRUE,results=TRUE}
baseball_lag <-  baseball %>%
  group_by(team) %>%
  mutate(diff_lag = log(payroll) - log(lag(payroll))) %>%
  drop_na() %>%
  select(team, year, diff_lag, win_pct)
head(baseball_lag)
```

## Exploratory questions

i. Which five teams had highest increase in their payroll between years 2010 and 2014, inclusive?

```{r, include=TRUE,echo=TRUE,results=TRUE}
baseball_lag %>% 
  filter(year >= 2010 & year <= 2014) %>% 
  arrange(desc(diff_lag)) %>% 
  head(5)
```

ii. Between 2010 and 2014, inclusive, which team(s) "improved" the most? That is, had the biggest percentage gain in wins?

```{r, include=TRUE,echo=TRUE,results=TRUE}
baseball_lag %>% 
  filter( year >= 2010 & year <= 2014) %>% 
  arrange(desc(win_pct)) %>% 
  head(5)
```


## Do log increases in payroll imply better performance? 

Is there evidence to support the hypothesis that higher increases in payroll on the log scale lead to increased performance?

Pick up a few statistics, accompanied with some data visualization, to support your answer.

```{r, include=TRUE,echo=TRUE}
baseball_lag %>%
  ggplot(aes(x = diff_lag, y = win_pct)) +
  geom_point(size = 2,alpha=.7) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = F,
    color = "red") +
  theme_bw() +
  labs(x = "Yearly Increase in Payroll (log)",
       y = "Winning Percentage",
       title = "Relationship between Yearly Increase in Payroll and Winning Percentage")

```

```{r, include=TRUE,echo=TRUE}
# OLS to model the relationship between payroll increase and winning percentage
model1 <- lm(win_pct ~diff_lag, data=baseball_lag)
# OLS model with year fixed effects
model1_FE <- lm(win_pct ~diff_lag+as.factor(year), data=baseball_lag)
#library(stargazer) - for regression table
#stargazer(model1, model1_FE,style = "apsr")
```

\begin{table}[!htbp] \centering 
  \caption{Relationship Between Yearly Payroll and Winning Percentage} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lcc} 
\\[-1.8ex]\hline \\[-1.8ex] 
\\[-1.8ex] & \multicolumn{2}{c}{Winning Percentage} \\ 
\\[-1.8ex] & (1) OLS & (2) OLS FE\\ 
\hline \\[-1.8ex] 
 Yearly Increase in Payroll (log) & 0.047$^{***}$ & 0.050$^{***}$  \\ 
  & (0.013) & (0.013)   \\ 
   &  &  \\ 
  Constant & 0.497$^{***}$ & 0.492$^{***}$ \\ 
  & (0.003) & (0.013) \\ 
  Year Fixed-Effects & No & Yes \\ 
 N & 480 & 480 \\ 
R$^{2}$ & 0.028 & 0.030 \\ 
Adjusted R$^{2}$ & 0.026 & $-$0.004 \\ 
Residual Std. Error & 0.070 (df = 478) & 0.071 (df = 463) \\ 
F Statistic & 13.700$^{***}$ (df = 1; 478) & 0.885 (df = 16; 463) \\ 
\hline \\[-1.8ex] 
\multicolumn{3}{l}{$^{*}$p $<$ .1; $^{**}$p $<$ .05; $^{***}$p $<$ .01} \\ 
\end{tabular} 
\end{table} 

Looking at Table 1 and the above scatterplot, there is evidence to support the hypothesis that higher increases in payroll on the log scale leads to increased performance. The scatterplot with a line of best fit shows a positive correlation between change in payroll and winning percentage. Moreover, using linear regression with ordinary least squares, we show that a 1-unit increase in logged yearly change in payroll is associated with a 0.047 increase in winning percentage. This relationship is robust to including year fixed effects.

## Comparison

Which set of factors are better explaining performance? Yearly payroll or yearly increase in payroll? What criterion is being used? 

```{r,include=TRUE,echo=TRUE,results=TRUE}
# Make predictions using the OLS model with yearly changes in payroll
model1_pred <- model1 %>% 
  predict(baseball_lag)
# OLS to model the relationship between payroll and winning percentage
model2 <- lm(win_pct ~payroll, data=baseball)
# Make predictions using the OLS model with yearly payroll
model2_pred <- model2 %>% 
  predict(baseball)

# Calculate RMSE for each model
mean((baseball_lag$win_pct - model1_pred)^2) %>% sqrt()
mean((baseball$win_pct - model2_pred)^2) %>% sqrt()

model1_summary <- summary(model1)
model1_summary[["r.squared"]]
model2_summary <- summary(model2)
model2_summary[["r.squared"]]

```

Using root mean square error as an evaluation metric, it looks like yearly payroll is a slightly better fit for the data than yearly increase in payroll. Using R-squared as an evaluation metric shows that yearly payroll explains more variance in the data than yearly increase in payroll. From both of these metrics, it looks like yearly payroll better explains performance than yearly increase in payroll.





