---
title: "Modern Data Mining, HW 3"
author:
- Ludwig Zhao
- Nick Pangakis
- Xinyi Wang
date: 'Due: 11:59Pm,  2/27, 2022'
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '4'
  html_document:
    code_folding: hide
    highlight: haddock
    number_sections: yes
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '4'
urlcolor: blue
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output
# check if you have ISLR package, if not, install it
if(!require('pacman')) {install.packages('pacman')}
pacman::p_load(ISLR, readxl, magrittr, dplyr, ggplot2,car) # add the packages needed
if(!require('leaps')) {install.packages('leaps')}
library('leaps')
```


\pagebreak

# Overview

Multiple regression is one of the most popular methods used in statistics as well as in machine learning. We use linear models as a working model for its simplicity and interpretability. It is important that we use domain knowledge as much as we could to determine the form of the response as well as the function format for the factors. Then, when we have many possible features to be included in the working model it is inevitable that we need to choose a best possible model with a sensible criterion. `Cp`, `BIC` and regularizations such as LASSO are introduced. Be aware that if a model selection is done formally or informally, the inferences obtained with the final `lm()` fit may not be valid. Some adjustment will be needed. This last step is beyond the scope of this class. Check the current research line that Linda and collaborators are working on. 

This homework consists of two parts: the first one is an excercise (you will feel it being a toy example after the covid case study) to get familiar with model selection skills such as, `Cp` and `BIC`. The main job is a rather involved case study about devastating covid19 pandemic.  Please read through the case study first. It is time that group members work together to run a real project. This project is for sure a great one listed in your CV. 

For covid case study, the major time and effort would be needed in EDA portion.

## Objectives

- Model building process

- Methods
    - Model selection
        + All subsets
        + Forward/Backward
    - Regularization
        + LASSO (L1 penalty)
        + Ridge (L2 penalty)
        + Elastic net
- Understand the criteria 
    - `Cp`
    - Testing Errors
    - `BIC` 
    - `K fold Cross Validation`
    - `LASSO` 
- Packages
    - `lm()`, `Anova`
    - `regsubsets()`
    - `glmnet()` & `cv.glmnet()`

# Review materials

- Study lecture: Model selection
- Study lecture: Regularization
- Study lecture: Multiple regression

Review the code and concepts covered during lectures: multiple regression, model selection and penalized regression through elastic net. 

# Case study 1:  `ISLR::Auto` data

This will be the last part of the Auto data from ISLR. The original data contains 408 observations about cars. It has some similarity as the Cars data that we use in our lectures. To get the data, first install the package `ISLR`. The data set `Auto` should be loaded automatically. We use this case to go through methods learned so far. 

Final modelling question: We want to explore the effects of each feature as best as possible. 

1) Preparing variables: 

a) You may explore the possibility of variable transformations. We normally do not suggest to transform $x$ for the purpose of interpretation. You may consider to transform $y$ to either correct the violation of the linear model assumptions or if you feel a transformation of $y$ makes more sense from some theory. In this case we suggest you to look into `GPM=1/MPG`. Compare residual plots of MPG or GPM as responses and see which one might yield a more satisfactory patterns. 

In addition, can you provide some background knowledge to support the notion: it makes more sense to model `GPM`?  

From comparing the residual plots, we can see that GPM (1/MPG) is a better response variable than GPM. The GPM residual plot shows a fairly random pattern and the estimated regression line is straight, indicating that the model is appropriate for the data. The MPG residual plot has a slight U-shape pattern and the estimated regression line is not straight, indicating that the model is less appropriate for the data. 

Fuel efficiencies are curvilinear, the higher mpg ratings go up, after about 20 mpg, the more efficiencies flatten out. The gpm, on the other hand, measures the gallon used per mile. 
```{r}
library(ISLR)
auto <- ISLR::Auto

auto$gpm <- 1 / auto$mpg
plot(lm(mpg ~ cylinders + displacement + horsepower + weight, data = auto), 1)
plot(lm(gpm ~ cylinders + displacement + horsepower + weight, data = auto), 1)

```

b) You may also explore by adding interactions and higher order terms. The model(s) should be as *parsimonious* (simple) as possible, unless the gain in accuracy is significant from your point of view. 

```{r}
auto$displacement_sqr <- auto$displacement^2

m1 <- lm(gpm ~ cylinders + displacement + horsepower + weight, data = auto)
summary(m1)
#fit with higher order tersm
m2 <- lm(gpm ~ cylinders + displacement + horsepower + weight + year, data = auto)
summary(m2)
m3 <- lm(gpm ~ cylinders + displacement + horsepower + weight + year + horsepower * year, data = auto)
summary(m3)

anova(m2,m3)

m4 <- lm(gpm ~ cylinders + displacement_sqr + horsepower + weight + year, data = auto)
summary(m4)

m5 <- lm(gpm ~ cylinders + displacement_sqr + horsepower + weight + year + horsepower * year, data = auto)
summary(m5)

anova(m4,m5)
```

c) Use Mallow's $C_p$ or BIC to select the model.

```{r}
#use Mallow's Cp to select the model
fit1 <- regsubsets(gpm ~ cylinders + displacement + horsepower + weight +
                     acceleration + year + origin, auto, nvmax = 7, method = "exhaustive")

f1 <- summary(fit1)
f1$which
f1$cp

plot(f1$cp, xlab = "Number of predictors", ylab = "Cp",
     col = "green", pch = 16)
#Cp shows that a model with 3 variables has smaller prediction error
#output the best model with 3 variables
coef(fit1, 3)
#look for the optimal model size by Cp
opt.size <- which.min(f1$cp)
opt.size
#look for optimal variables selected
fit1.var <- f1$which
fit1.var[opt.size,]

colnames(fit1.var)[fit1.var[3,]]
```


2) Describe the final model and its accuracy. Include diagnostic plots with particular focus on the model residuals.

  * Summarize the effects found.
  * Predict the `mpg` of a car that is: built in 1983, in the US, red, 180 inches long, 8 cylinders, 350 displacement, 260 as horsepower, and weighs 4,000 pounds. Give a 95% CI.
  * Any suggestions as to how to improve the quality of the study?


```{r}
fit.final <- lm(gpm ~ horsepower + weight + year, data=auto)
summary(fit.final)
plot(fit.final,1:2)
Anova(fit.final)

newData <- data.frame(year = 83, cylinders = 8, displacement = 350, weight = 4000, horsepower = 260, origin = 1)
fit.new <- predict(fit.final, newData, interval = "confidence", level = 0.95)
1/fit.new
```

The final model uses horsepower, weight and year as the predictors. 
The residual sum of square is 0.01289 and $R^2$ is 0.881. All variables are statically significant.
From the regression, the ``gpm`` increases as ``horsepower`` and ``weight`` increase. 
``gpm`` decreases as ``year`` increase. 

For the prediction, the new car is expected to have ``mpg`` of 15.6 with 95% CI from 14.6 to 16.7.

For the suggestion to improve the study, one way is to exclude those cars that are outlines, such as those sports cars as they may have different ``mpg`` prediction model compare to regular cars.

# Case study 2: COVID

See covid_case_study.Rmd.

